---
title: "CMS Medicare Physician & Other Practitioners by Provider dataset for 2023 research questions"
subtitle: "INFO 6105 Final Project"
author: "Dany De Leriol Nohonet Sigha"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup-preprocessing, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

```{=html}
<style>
body { font-family: 'Palatino', 'Georgia', serif; }
h1, h2 { color: #C8102E; }
h3 { color: #444; }
.try-it {
  border: 2px solid #C8102E; padding: 12px; margin: 10px 0;
  background-color: #fdf5f5; border-radius: 5px;
}
.why-box {
  border-left: 5px solid #2166AC; padding: 10px 15px;
  margin: 10px 0; background-color: #f0f6fc;
}
.pattern-box {
  border: 2px solid #28A745; padding: 12px; margin: 10px 0;
  background-color: #f0fff4; border-radius: 5px;
}
</style>
```

------------------------------------------------------------------------

# Welcome

This is the data mdeling step for the "Medicare Payment Gap Analysis in Massachussets" final project. 

This file will answer the research questions (RQ):

- RQ1: Perform multiple linear regression including coefficient interpretation (“holding all else constant”), 

- RQ1: significance tests, R2 and adjusted R2, VIF for multicollinearity, and diagnostic plots (residuals vs. fitted, Q-Q plot, scale-location plot). Check LINE as-sumptions

- RQ2: Perform One-Way ANOVA to compare payment gap means accross provider specialties

- RQ2: State hypotheses, conduct F-test, create comparison plots, perform Tukey HSD post-hoc tests, check assumptions (normality, equal variances), and calculate effect size (η2)

- RQ3: Model a binary outcome, the log-odds of being a high-gap provider.

- RQ3: Report odds ratios with interpretations (“holding other variables constant”), confusion matrix with accuracy/sensitivity/specificity/precision, and ROC curve with AUC. Discuss threshold choice.

- RQ4: Use bootstrap to construct a CI for the median payment gap between medical providers of high pct_dual and low_pct_dual. 

- RQ4: Use a permutation test to test a hypothesis. Compare computational results to parametric results.

**Dataset:** `MUP_PHY_R25_P05_V20_D23_Prov.csv` — 1,259,343 rows, each a provider--procedure--setting combination in 2023. Each row represents the aggregated activity for a specific provider performing a specific procedure (identified by HCPCS code) in a specific care setting (facility or non-facility) over the full 2023 reporting year, capturing totals such as number of times the service was rendered, number of unique beneficiaries who received it, total submitted charges, and total Medicare payments for that provider-procedure-setting combination.

Missing values are absent across these key variables, including payment_gap, Tot_Sbmtd_Chrg, Tot_Mdcr_Stdzd_Amt, Tot_Benes, Tot_Srvcs, Rndrng_Prvdr_Type, Bene_Avg_Age, Bene_Avg_Risk_Scre. However, some chronic condition percentage variables are missing, and the Drug_Sbmtd_Chrg is sometimes absent.

------------------------------------------------------------------------

# Setup

```{r preprocessing-setup}
# Install if needed (uncomment once, then re-comment)

library(tidyverse)
library(haven)
library(broom)
library(car)
library(pROC)
library(boot)
library(multcomp)
library(rstatix)
library(multcompView)
library(parameters)
library(quantreg)
library(dunn.test)
library(ggrepel)

# Explicitly prefer dplyr functions over masked alternatives
select    <- dplyr::select
filter    <- dplyr::filter
summarise <- dplyr::summarise

theme_set(theme_minimal(base_size = 13))
```

------------------------------------------------------------------------

# Part 1: Multiple Linear Regression (RQ1)

::: fill-box
**Research Question 1:** Do provider volume, patient risk profile, and chronic disease
burden predict the standardized Medicare payment gap among Massachusetts providers, and
which factors have the strongest independent effect holding all others constant?

**Why multiple linear regression?** The outcome variable `log(payment_gap)` is
continuous. MLR models the outcome as a linear combination of predictors, allowing each
coefficient to be interpreted as the expected change in log payment gap for a one-unit
increase in that predictor, holding all others constant. Exponentiated coefficients give
multiplicative effects on the raw payment gap.

**Two-model structure:** Model 1 uses the full dataset (n = 37,708) with risk score and
age as the sole complexity controls, since `Bene_CC_*` variables are suppressed for 90%
of providers. Model 2 uses the complete-CC subset (n = 3,752) and adds chronic condition
variables selected by LASSO. The two models are complementary --- Model 1 establishes
baseline drivers across all Massachusetts providers, Model 2 tests whether specific
chronic condition burden adds explanatory power among high-volume providers where it can
be reliably measured.

**Predictors:**

*Model 1 (full dataset):*
- `Tot_Benes` — provider volume (practice size)
- `srvcs_per_bene` — billing intensity (services per beneficiary)
- `Bene_Avg_Age` — average beneficiary age
- `Bene_Avg_Risk_Scre` — average HCC risk score (composite patient complexity)

*Model 2 (complete-CC subset):*
- All Model 1 predictors, plus LASSO-selected chronic condition variables:
  `Bene_CC_PH_Hyperlipidemia_V2_Pct`, `Bene_CC_PH_Hypertension_V2_Pct`,
  `Bene_CC_BH_PD_V1_Pct`, `Bene_CC_PH_Asthma_V2_Pct`,
  `Bene_CC_BH_Mood_V2_Pct`, `Bene_CC_PH_Arthritis_V2_Pct`,
  `Bene_CC_BH_Alz_NonAlzdem_V2_Pct`, `Bene_CC_BH_Alcohol_Drug_V1_Pct`,
  `Bene_CC_PH_HF_NonIHD_V2_Pct`
:::

------------------------------------------------------------------------

## Model 1: Full dataset

```{r p1-model1-fit}
ma_providers <- readRDS("../data/processed/ma_providers_preprocessed.rds")
ma_providers_rq3 <- readRDS("../data/processed/ma_providers_rq3.rds")

#model1 <- lm(
#  log(payment_gap) ~ Tot_Benes + srvcs_per_bene + Bene_Avg_Age + Bene_Avg_Risk_Scre,
#  data = ma_providers
#)

model1 <- lm(
  log(payment_gap) ~ log(Tot_Benes) + log(srvcs_per_bene) + Bene_Avg_Age + Bene_Avg_Risk_Scre,
  data = ma_providers
)

cat("── Model 1 Summary ──\n")
summary(model1)


```


```{r p1-model1-tidy}
cat("── Model 1 Coefficients (tidy) ──\n")
tidy(model1, conf.int = TRUE) %>%
  mutate(
    exp_estimate = exp(estimate),
    exp_conf.low  = exp(conf.low),
    exp_conf.high = exp(conf.high)
  ) %>%
  print()
```

::: pattern-box
**Interpreting Model 1 coefficients:**

Coefficients are on the log scale. Exponentiating gives the multiplicative effect on the
raw payment gap for a one-unit increase in each predictor. For example, if the coefficient
for `srvcs_per_bene` is 0.15, then each additional service per beneficiary is associated
with a payment gap that is $e^{0.15} \approx 1.16$ times larger — a 16% increase —
holding all other predictors constant.
:::

```{r p1-model1-fit-stats}
cat("── Model 1 R² and Adjusted R² ──\n")
cat("R²:         ", round(summary(model1)$r.squared, 4), "\n")
cat("Adjusted R²:", round(summary(model1)$adj.r.squared, 4), "\n")

cat("\n── Model 1 VIF (multicollinearity check) ──\n")
vif(model1)
```

::: why-box
**Variance Inflation Factor (VIF):**

VIF quantifies how much the variance of a coefficient is inflated due to collinearity
with other predictors. A VIF above 5 indicates moderate multicollinearity and above 10
indicates severe multicollinearity that may destabilize coefficient estimates. Because
`Bene_Avg_Risk_Scre` is a composite that already incorporates chronic condition burden,
mild correlation with `Bene_Avg_Age` is expected and acceptable.
:::

```{r p1-model1-diagnostics, fig.cap="Model 1 diagnostic plots. Residuals vs Fitted assesses linearity and homoscedasticity; Q-Q plot assesses normality of residuals; Scale-Location plot assesses equal variance; Residuals vs Leverage identifies influential observations."}
par(mfrow = c(2, 2))
plot(model1)
par(mfrow = c(1, 1))
```

```{r p1-model1-cooks}
cat("── Model 1 influential observations (Cook's distance > 4/n) ──\n")
cooks_d <- cooks.distance(model1)
threshold <- 4 / nrow(ma_providers)
influential <- which(cooks_d > threshold)
cat("Number of influential observations:", length(influential), "\n")
cat("Proportion of dataset:", round(length(influential) / nrow(ma_providers) * 100, 1), "%\n")
```

::: why-box
**Why complement Model 1 with quantile regression?**
OLS regression estimates a single coefficient per predictor — the
average effect across all 37,708 Massachusetts providers. This average
is meaningful only if each predictor's effect is roughly uniform across
the payment gap distribution. Three of four LINE assumptions are
violated in Model 1, and the diagnostic plots show that residual spread
is largest among low-volume providers and shrinks at high volume — a
pattern consistent with heterogeneous predictor effects across the
distribution.

Quantile regression addresses this directly. Rather than targeting the
conditional mean, it targets a specified quantile of the conditional
outcome distribution — here the 25th, 50th, and 75th percentiles.
This allows each predictor's coefficient to vary across the gap
distribution, revealing whether billing intensity, patient complexity,
or volume matter differently for low-gap versus high-gap providers.
No distributional assumptions are required, making the approach robust
to the heteroscedasticity and heavy tails documented above.

The same four predictors as Model 1 are used:
\texttt{log(Tot\_Benes)}, \texttt{log(srvcs\_per\_bene)},
\texttt{Bene\_Avg\_Age}, and \texttt{Bene\_Avg\_Risk\_Scre}.
Standard errors are estimated via the kernel method (\texttt{se = "ker"}),
which is well suited for large continuous outcomes and avoids the
instability of sandwich estimators at $n = 37{,}708$.
:::

```{r p1-model1-quantile}
qr_model1 <- rq(
  log(payment_gap) ~ log(Tot_Benes) + log(srvcs_per_bene) + Bene_Avg_Age + Bene_Avg_Risk_Scre,
  tau  = c(0.25, 0.50, 0.75),
  data = ma_providers
)

cat("── Quantile Regression Summary (Q25, Q50, Q75) ──\n")
summary(qr_model1, se = "ker")
```

::: fill-box
**Reading the quantile regression coefficient plot:**
Each panel in the figure below shows one predictor. The x-axis indexes
the quantile of the conditional payment gap distribution (Q25, Q50,
Q75). The blue points are the quantile regression coefficient estimates
at each quantile, with 95% confidence intervals shown as vertical bars.
The orange dotted line is the OLS coefficient from Model 1 — the single
average estimate. The orange shaded band is the OLS 95% confidence
interval.

**What to look for:**

- If the blue points are flat and fall inside the orange band, OLS is
  adequate for that predictor — its effect is uniform across the
  distribution.
- If the blue points trend upward from Q25 to Q75 and stray outside
  the orange band, the predictor has a stronger effect on high-gap
  providers than low-gap ones — OLS understates it at the top and
  overstates it at the bottom.
- If the blue points trend downward, the predictor matters more for
  low-gap providers — OLS overstates its effect at the top of the
  distribution.
- A sign change across quantiles indicates the predictor has opposite
  associations at different parts of the distribution.
:::

```{r p1-model1-quantile-plot}

# Extract quantile regression coefficients at each tau
qr_tidy <- bind_rows(
  lapply(c(0.25, 0.50, 0.75), function(tau) {
    fit <- rq(
      log(payment_gap) ~ log(Tot_Benes) + log(srvcs_per_bene) +
        Bene_Avg_Age + Bene_Avg_Risk_Scre,
      tau  = tau,
      data = ma_providers
    )
    s        <- summary(fit, se = "ker")
    coef_mat <- s$coefficients
    data.frame(
      term     = rownames(coef_mat),
      estimate = coef_mat[, "Value"],
      se       = coef_mat[, "Std. Error"],
      tau      = tau
    )
  })
) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    lower     = estimate - 1.96 * se,
    upper     = estimate + 1.96 * se,
    tau_label = paste0("Q", tau * 100)
  )

# OLS reference
ols_tidy <- tidy(model1, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  select(term, estimate, conf.low, conf.high)

# Plot
ggplot() +
  # OLS CI shaded rectangle per facet
  geom_rect(
    data = ols_tidy,
    aes(
      xmin = 0.20,
      xmax = 0.80,
      ymin = conf.low,
      ymax = conf.high
    ),
    fill  = "#E66101",
    alpha = 0.15
  ) +
  # OLS point estimate dotted line
  geom_hline(
    data      = ols_tidy,
    aes(yintercept = estimate),
    colour    = "#E66101",
    linetype  = "dotted",
    linewidth = 0.8
  ) +
  # Quantile regression line
  geom_line(
    data      = qr_tidy,
    aes(x = tau, y = estimate),
    colour    = "#2166AC",
    linewidth = 0.8
  ) +
  # Quantile regression points with CI
  geom_pointrange(
    data = qr_tidy,
    aes(
      x    = tau,
      y    = estimate,
      ymin = lower,
      ymax = upper
    ),
    colour = "#2166AC",
    size   = 0.5
  ) +
  facet_wrap(~ term, scales = "free_y", ncol = 2) +
  scale_x_continuous(
    breaks = c(0.25, 0.50, 0.75),
    labels = c("Q25", "Q50", "Q75"),
    limits = c(0.20, 0.80)
  ) +
  labs(
    title    = "Quantile Regression Coefficients vs OLS — Model 1",
    subtitle = "Orange band/line = OLS estimate and 95% CI; blue = quantile regression at Q25, Q50, Q75",
    x        = "Quantile",
    y        = "Coefficient estimate"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13),
    plot.subtitle = element_text(size = 10, colour = "grey40"),
    strip.text    = element_text(size = 10, face = "bold")
  )

ggsave(
  here::here("docs", "figures", "quantile_coef_model1.png"),
  width = 9, height = 6, dpi = 300
)
```

::: pattern-box
**What the quantile regression reveals:**
Four findings emerge that OLS cannot produce.

**log(Tot\_Benes)** is the most stable predictor across quantiles
(Q25: 0.990, Q50: 0.997, Q75: 0.997), but all three estimates lie
above the OLS estimate of 0.971 and outside its confidence band.
Volume affects all providers similarly regardless of where they sit
in the gap distribution — but OLS slightly underestimates this effect.

**log(srvcs\_per\_bene)** declines monotonically from 0.477 at Q25
to 0.375 at Q75, with all three estimates outside the OLS band of
0.417. Billing intensity is a stronger lever for low-gap providers
than for high-gap ones. Among providers already at the top of the
distribution, billing more intensively produces a smaller marginal
increase in the gap than OLS predicts.

**Bene\_Avg\_Risk\_Scre** shows the steepest cross-quantile gradient,
declining from 0.311 at Q25 to 0.171 at Q75. Patient complexity drives
the gap meaningfully among low-gap providers but contributes relatively
little among high-gap ones — those providers appear to bill aggressively
regardless of how sick their patient panels are.

**Bene\_Avg\_Age** changes sign across quantiles, from positive at Q25
(0.003) to negative at Q75 ($-$0.001), and is non-significant at Q75
($p = 0.251$). This sign reversal confirms that age is not a stable
predictor across the distribution, consistent with its non-significance
in the OLS model.

Taken together: the predictors that drive the gap among low-gap
providers — billing intensity and patient complexity — lose explanatory
power at the top of the distribution, where volume dominates. OLS was
averaging across this heterogeneity and missing it entirely.
:::

------------------------------------------------------------------------

## Model 2: Complete-CC subset

```{r p1-model2-fit}
ma_providers_cc <- ma_providers %>%
  filter(!any_cc_missing)

cc_predictors <- c(
  "Bene_CC_PH_Hyperlipidemia_V2_Pct",
  "Bene_CC_PH_Hypertension_V2_Pct",
  "Bene_CC_BH_PD_V1_Pct",
  "Bene_CC_PH_Asthma_V2_Pct",
  "Bene_CC_PH_Arthritis_V2_Pct",
  "Bene_CC_BH_Alz_NonAlzdem_V2_Pct",
  "Bene_CC_BH_Alcohol_Drug_V1_Pct",
  "Bene_CC_PH_HF_NonIHD_V2_Pct"
)

#model2_formula <- as.formula(
#  paste(
#    "log(payment_gap) ~ Tot_Benes + srvcs_per_bene + Bene_Avg_Age + Bene_Avg_Risk_Scre +",
#    paste(cc_predictors, collapse = " + ")
#  )
#)

model2_formula <- as.formula(
  paste(
    "log(payment_gap) ~ log(Tot_Benes) + log(srvcs_per_bene) + Bene_Avg_Age + Bene_Avg_Risk_Scre +",
    paste(cc_predictors, collapse = " + ")
  )
)

model2 <- lm(model2_formula, data = ma_providers_cc)

cat("── Model 2 Summary ──\n")
summary(model2)
```

```{r p1-model2-tidy}
cat("── Model 2 Coefficients (tidy) ──\n")
tidy(model2, conf.int = TRUE) %>%
  mutate(
    exp_estimate  = exp(estimate),
    exp_conf.low  = exp(conf.low),
    exp_conf.high = exp(conf.high)
  ) %>%
  print()
```

```{r p1-model2-fit-stats}
cat("── Model 2 R² and Adjusted R² ──\n")
cat("R²:         ", round(summary(model2)$r.squared, 4), "\n")
cat("Adjusted R²:", round(summary(model2)$adj.r.squared, 4), "\n")

cat("\n── Model 2 VIF (multicollinearity check) ──\n")
vif(model2)
```

```{r p1-model2-diagnostics, fig.cap="Model 2 diagnostic plots for the complete-CC subset."}
par(mfrow = c(2, 2))
plot(model2)
par(mfrow = c(1, 1))
```

```{r p1-model2-cooks}
cat("── Model 2 influential observations (Cook's distance > 4/n) ──\n")
cooks_d2 <- cooks.distance(model2)
threshold2 <- 4 / nrow(ma_providers_cc)
influential2 <- which(cooks_d2 > threshold2)
cat("Number of influential observations:", length(influential2), "\n")
cat("Proportion of dataset:", round(length(influential2) / nrow(ma_providers_cc) * 100, 1), "%\n")
```

::: why-box
**Why extend Model 2 with quantile regression?**
Model 2 already operates on a non-random subset of the Massachusetts
provider population --- the 3,752 high-volume providers for whom CMS
reports complete chronic condition data. These providers have a median
payment gap of \$404,539, more than four times the full-population
median of \$89,292. OLS estimates a single average effect for each of
the 12 predictors within this already-elevated stratum, but that
average may not represent any particular provider well if the effects
are heterogeneous within the subset.

The same LINE violations documented in Model 1 apply here, and the
smaller sample size ($n = 3{,}752$) means the asymptotic defence is
somewhat weaker. More importantly, the CC-complete subset has a
substantively different composition from the full population --- it is
dominated by high-volume procedural specialists whose billing behaviour
may differ systematically from the median provider. If billing
intensity or chronic condition effects operate differently among
over-billers versus under-billers within this stratum, OLS will average
across that heterogeneity and potentially misrepresent both groups.

Quantile regression at Q25, Q50, and Q75 addresses this directly.
Within the CC-complete subset, Q25 represents under-billers relative
to their profile among high-volume providers, Q50 represents typical
high-volume billers, and Q75 represents over-billers. Because the
entire subset is already in the upper tier of the full gap distribution,
these quantiles characterise heterogeneity within an elevated stratum
--- a finer-grained view than Model 1 could provide.
:::

```{r p1-model2-quantile}

qr_model2_formula <- as.formula(
  paste(
    "log(payment_gap) ~ log(Tot_Benes) + log(srvcs_per_bene) + Bene_Avg_Age + Bene_Avg_Risk_Scre +",
    paste(cc_predictors, collapse = " + ")
  )
)

qr_model2 <- rq(
  qr_model2_formula,
  tau  = c(0.25, 0.50, 0.75),
  data = ma_providers_cc
)

cat("── Quantile Regression Summary Model 2 (Q25, Q50, Q75) ──\n")
summary(qr_model2, se = "ker")
```

::: fill-box
**Reading the Model 2 quantile coefficient plots:**
Model 2 produces two coefficient plots --- one for the four base
predictors shared with Model 1, and one for the eight LASSO-selected
chronic condition predictors. Both follow the same visual logic as the
Model 1 plot: blue points and lines show quantile regression estimates
at Q25, Q50, and Q75 with 95\% confidence intervals; the orange dotted
line is the OLS coefficient; the orange shaded band is the OLS 95\%
confidence interval.

**For the base predictors plot**, the comparison to Model 1 is
informative. If the quantile patterns here differ from Model 1 ---
for instance, if billing intensity now increases rather than decreases
across quantiles --- that signals that the CC-complete subset is a
genuinely different population where predictors operate differently,
not just a smaller version of the full dataset.

**For the chronic condition predictors plot**, look for two things.
First, which conditions have stable effects across quantiles (points
flat and inside the OLS band) versus heterogeneous effects (points
trending up or down and straying outside the band). Stable conditions
are reliable predictors for all high-volume providers; heterogeneous
ones matter more for specific tiers. Second, pay attention to whether
any conditions that appeared strong in OLS turn out to be concentrated
at specific quantiles --- a large OLS coefficient driven entirely by
Q25 or Q75 effects tells a very different policy story than one that
is uniform across the distribution.
:::

```{r p1-model2-quantile-plot}

# Helper to clean up term labels
recode_terms <- function(x) {
  case_when(
    x == "log(Tot_Benes)"                   ~ "log(Tot_Benes)",
    x == "log(srvcs_per_bene)"              ~ "log(srvcs_per_bene)",
    x == "Bene_Avg_Age"                      ~ "Bene_Avg_Age",
    x == "Bene_Avg_Risk_Scre"               ~ "Bene_Avg_Risk_Scre",
    x == "Bene_CC_PH_Hyperlipidemia_V2_Pct" ~ "Hyperlipidemia",
    x == "Bene_CC_PH_Hypertension_V2_Pct"   ~ "Hypertension",
    x == "Bene_CC_BH_PD_V1_Pct"            ~ "Personality Disorder",
    x == "Bene_CC_PH_Asthma_V2_Pct"        ~ "Asthma",
    x == "Bene_CC_PH_Arthritis_V2_Pct"      ~ "Arthritis",
    x == "Bene_CC_BH_Alz_NonAlzdem_V2_Pct" ~ "Dementia",
    x == "Bene_CC_BH_Alcohol_Drug_V1_Pct"  ~ "Alcohol/Drug",
    x == "Bene_CC_PH_HF_NonIHD_V2_Pct"     ~ "Heart Failure",
    TRUE ~ x
  )
}

qr_tidy2 <- bind_rows(
  lapply(c(0.25, 0.50, 0.75), function(tau) {
    fit <- rq(
      qr_model2_formula,
      tau  = tau,
      data = ma_providers_cc
    )
    s        <- summary(fit, se = "ker")
    coef_mat <- s$coefficients
    data.frame(
      term     = rownames(coef_mat),
      estimate = coef_mat[, "Value"],
      se       = coef_mat[, "Std. Error"],
      tau      = tau
    )
  })
) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    lower     = estimate - 1.96 * se,
    upper     = estimate + 1.96 * se,
    tau_label = paste0("Q", tau * 100),
    term      = recode_terms(term)
  )

# OLS reference from model2
ols_tidy2 <- tidy(model2, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = recode_terms(term)) %>%
  select(term, estimate, conf.low, conf.high)

# Split into base and CC predictors
base_terms <- c("log(Tot_Benes)", "log(srvcs_per_bene)",
                 "Bene_Avg_Age", "Bene_Avg_Risk_Scre")

cc_terms <- c("Hyperlipidemia", "Hypertension", "Personality Disorder",
               "Asthma", "Arthritis", "Dementia", "Alcohol/Drug", "Heart Failure")

plot_qr <- function(tidy_qr, tidy_ols, terms, title_suffix) {
  ggplot() +
    geom_rect(
      data = tidy_ols %>% filter(term %in% terms),
      aes(
        xmin = 0.20, xmax = 0.80,
        ymin = conf.low, ymax = conf.high
      ),
      fill  = "#E66101",
      alpha = 0.15
    ) +
    geom_hline(
      data      = tidy_ols %>% filter(term %in% terms),
      aes(yintercept = estimate),
      colour    = "#E66101",
      linetype  = "dotted",
      linewidth = 0.8
    ) +
    geom_line(
      data      = tidy_qr %>% filter(term %in% terms),
      aes(x = tau, y = estimate),
      colour    = "#2166AC",
      linewidth = 0.8
    ) +
    geom_pointrange(
      data = tidy_qr %>% filter(term %in% terms),
      aes(x = tau, y = estimate, ymin = lower, ymax = upper),
      colour = "#2166AC",
      size   = 0.5
    ) +
    facet_wrap(~ term, scales = "free_y", ncol = 2) +
    scale_x_continuous(
      breaks = c(0.25, 0.50, 0.75),
      labels = c("Q25", "Q50", "Q75"),
      limits = c(0.20, 0.80)
    ) +
    labs(
      title    = paste("Quantile Regression Coefficients vs OLS — Model 2,", title_suffix),
      subtitle = "Orange band/line = OLS estimate and 95% CI; blue = quantile regression at Q25, Q50, Q75",
      x        = "Quantile",
      y        = "Coefficient estimate"
    ) +
    theme_minimal() +
    theme(
      plot.title    = element_text(size = 13),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      strip.text    = element_text(size = 10, face = "bold")
    )
}

p_base <- plot_qr(qr_tidy2, ols_tidy2, base_terms, "Base Predictors")
p_cc   <- plot_qr(qr_tidy2, ols_tidy2, cc_terms,   "Chronic Condition Predictors")

print(p_base)
print(p_cc)

ggsave(
  here::here("docs", "figures", "quantile_coef_model2_base.png"),
  plot  = p_base,
  width = 9, height = 6, dpi = 300
)

ggsave(
  here::here("docs", "figures", "quantile_coef_model2_cc.png"),
  plot  = p_cc,
  width = 9, height = 6, dpi = 300
)
```
::: pattern-box
**What the Model 2 quantile regression reveals:**

**Base predictors --- two key reversals from Model 1:**

**log(Tot\_Benes)** is flat across quantiles (Q25: 0.441, Q50: 0.420,
Q75: 0.462) but lies below the OLS estimate of 0.555 at all three
points, outside the OLS confidence band. Within the CC-complete
high-volume subset, OLS overestimates the volume effect. Volume has
already done much of its work by selecting providers into this
high-gap stratum; within-stratum variation in volume carries less
marginal explanatory power than OLS suggests.

**log(srvcs\_per\_bene)** reverses the Model 1 pattern, increasing
from 0.444 at Q25 to 0.580 at Q75, with Q75 outside the OLS band.
In Model 1, billing intensity mattered more for low-gap providers.
Here, within the high-volume subset, it matters more for over-billers.
This reversal is substantively important: aggressive billing intensity
is a distinguishing characteristic of over-billers specifically within
this stratum.

**Bene\_Avg\_Risk\_Scre** declines sharply from 0.297 at Q25 to 0.032
at Q75 (non-significant, $p = 0.548$). Among over-billers in this
high-volume subset, patient complexity adds almost nothing to the gap
once chronic condition mix is controlled for.

**Chronic condition predictors --- three findings worth highlighting:**

**Dementia** is the most consistent predictor across all quantiles
($-$0.021, $-$0.021, $-$0.018), all within the OLS band and all
significant. The gap-reducing effect of dementia prevalence is uniform
within this subset --- it applies equally to under-billers, typical
providers, and over-billers.

**Hypertension** shows an unexpected V-shape: 0.037 at Q25, dropping
to near zero at Q50 (0.009, non-significant), then recovering to 0.021
at Q75. The OLS coefficient of 0.052 --- the largest positive chronic
condition effect in the paper --- was masking this instability. The
hypertension effect is concentrated at the extremes of the distribution
and essentially absent for typical high-volume providers. This
substantially qualifies the headline OLS finding.

**Hyperlipidemia** declines from 0.048 at Q25 to 0.020 at Q75, with
Q25 above the OLS band. Its positive effect is concentrated among
under-billers within this stratum and attenuates for over-billers.

**Alcohol/Drug**, **Heart Failure**, and **Asthma** are broadly
consistent with OLS estimates and stable across quantiles.
**Personality Disorder** and **Arthritis** show modest attenuation
at higher quantiles but remain within their OLS confidence bands.
:::

------------------------------------------------------------------------

## Model comparison

```{r p1-model-comparison}
cat("── Model 1 vs Model 2 comparison ──\n")
bind_rows(
  glance(model1) %>% mutate(model = "Model 1 (full, n = 37,708)"),
  glance(model2) %>% mutate(model = "Model 2 (complete-CC, n = 3,752)")
) %>%
  dplyr::select(model, r.squared, adj.r.squared, AIC, BIC, statistic, p.value) %>%
  print()
```
::: why-box
**Why check coefficient stability after removing influential
observations?**
Cook's distance flags observations that have disproportionate
influence on the OLS coefficient estimates. Model 1 identifies
1,899 such observations (5.0\% of the dataset) using the
conventional threshold of $4/n$. These are providers whose
combination of predictor values and payment gap makes them
outliers relative to the fitted model.

The presence of influential observations is not itself a problem
--- large datasets always contain some. The question is whether
the coefficient estimates change materially when these observations
are excluded. If the estimates are stable, the findings are robust
and the influential observations are not driving the results. If
they change substantially, that signals that the OLS estimates
are sensitive to a small number of extreme providers, which would
be an important qualification of the main findings.

Model 1 is therefore refit on the 35,809 providers remaining
after removing the 1,899 influential observations, and the
coefficients are compared directly.
:::

::: fill-box
**Reading the Cook's distance robustness plot:**
Each panel shows one predictor. The left point in each panel is
the coefficient estimate from the full Model 1 ($n = 37{,}708$);
the right point is the coefficient from the model refit after
removing influential observations ($n = 35{,}809$). Vertical
bars show 95\% confidence intervals.

**What to look for:** if the two points in each panel overlap
or are very close relative to their confidence intervals, the
coefficient is stable --- the influential observations are not
driving that finding. If the points are far apart and their
confidence intervals do not overlap, the finding is sensitive
to the influential observations and should be interpreted with
more caution.

The R² comparison is also informative. An increase in R² after
removing influential observations indicates that those
observations were poorly fitted outliers pulling the model fit
down. A decrease would indicate the influential observations
were well-fitted high-leverage points helping the model.
:::

```{r p1-model1-cooks-robustness}
# Identify influential observations
cooks_d    <- cooks.distance(model1)
threshold  <- 4 / nrow(ma_providers)
influential_idx <- which(cooks_d > threshold)

cat("── Model 1 refitted without influential observations ──\n")
cat("Removing", length(influential_idx), "influential observations\n")
cat("Remaining n:", nrow(ma_providers) - length(influential_idx), "\n\n")

# Refit without influential observations
ma_providers_clean <- ma_providers[-influential_idx, ]

model1_clean <- lm(
  log(payment_gap) ~ log(Tot_Benes) + log(srvcs_per_bene) +
    Bene_Avg_Age + Bene_Avg_Risk_Scre,
  data = ma_providers_clean
)

cat("── Coefficient comparison: full vs cleaned dataset ──\n")
bind_rows(
  tidy(model1) %>%
    mutate(dataset = "Full (n = 37,708)"),
  tidy(model1_clean) %>%
    mutate(dataset = paste0("Cleaned (n = ",
                            nrow(ma_providers_clean), ")"))
) %>%
  filter(term != "(Intercept)") %>%
  select(dataset, term, estimate, std.error, p.value) %>%
  mutate(across(c(estimate, std.error), ~round(., 4))) %>%
  mutate(p.value = round(p.value, 6)) %>%
  arrange(term, dataset) %>%
  print(n = Inf)

cat("\n── R² comparison ──\n")
cat("Full model R²:   ", round(summary(model1)$r.squared, 4), "\n")
cat("Cleaned model R²:", round(summary(model1_clean)$r.squared, 4), "\n")
```

```{r p1-model1-cooks-robustness-plot}

# Pre-compute label strings to avoid paste0 inside aes/scale calls
label_full    <- "Full dataset\n(n = 37,708)"
label_cleaned <- paste0("Influential obs. removed\n(n = ",
                         nrow(ma_providers_clean), ")")

coef_comparison <- bind_rows(
  tidy(model1, conf.int = TRUE) %>%
    mutate(dataset = label_full),
  tidy(model1_clean, conf.int = TRUE) %>%
    mutate(dataset = label_cleaned)
) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "log(Tot_Benes)"      ~ "log(Tot_Benes)",
      term == "log(srvcs_per_bene)" ~ "log(srvcs_per_bene)",
      term == "Bene_Avg_Age"        ~ "Bene_Avg_Age",
      term == "Bene_Avg_Risk_Scre"  ~ "Bene_Avg_Risk_Scre",
      TRUE ~ term
    )
  )

ggplot(coef_comparison,
       aes(x = dataset, y = estimate,
           ymin = conf.low, ymax = conf.high,
           colour = dataset)) +
  geom_pointrange(size = 0.6) +
  facet_wrap(~ term, scales = "free_y", ncol = 2) +
  scale_colour_manual(
  values = setNames(
      c("#2166AC", "#C8102E"),
      c(label_full, label_cleaned)
    ),
    guide = "none"
  ) +
  labs(
    title    = "Model 1 Coefficient Stability — Cook's Distance Robustness Check",
    subtitle = "Blue = full dataset; Red = influential observations removed (Cook's D > 4/n)",
    x        = NULL,
    y        = "Coefficient estimate"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13),
    plot.subtitle = element_text(size = 10, colour = "grey40"),
    strip.text    = element_text(size = 10, face = "bold")
  )

ggsave(
  here::here("docs", "figures", "cooks_robustness_model1.png"),
  width = 8, height = 5, dpi = 300
)
```

::: pattern-box
**What the robustness check reveals:**
R² increases from 0.629 to 0.738 after removing the 1,899
influential observations, confirming that these providers were
poorly fitted outliers rather than high-leverage points
supporting the model fit. Their removal improves fit
substantially.

**\texttt{log(Tot\_Benes)}}** is the most stable predictor:
0.971 in the full model versus 0.998 in the cleaned model,
with overlapping confidence intervals. The dominant finding
--- that provider volume is the strongest predictor of the
payment gap --- does not depend on influential observations.

**\texttt{log(srvcs\_per\_bene)}}** is equally stable: 0.417
versus 0.406, a negligible difference.

**\texttt{Bene\_Avg\_Risk\_Scre}** shows the most meaningful
shift: 0.312 in the full model versus 0.256 in the cleaned
model, an 18\% attenuation. The risk score effect is partially
inflated by the influential observations. It remains
significant and in the same direction, but its magnitude
should be interpreted with some caution.

**\texttt{Bene\_Avg\_Age}** shows a substantively interesting
change: non-significant in the full model ($p = 0.213$) but
significant in the cleaned model (0.0023, $p = 0.002$). The
influential observations were suppressing the age effect.
This further supports its retention in the model --- its
non-significance in the full dataset reflects the distorting
influence of outliers rather than a genuine absence of effect.

Overall the robustness check confirms that the core Model 1
findings are not driven by influential observations. The one
qualification worth noting is that the patient complexity
effect (\texttt{Bene\_Avg\_Risk\_Scre}) is modestly inflated
in the full dataset, consistent with the quantile regression
finding that this predictor attenuates at higher quantiles
where extreme providers are more prevalent.
:::
------------------------------------------------------------------------

# Part 2: One-Way ANOVA (RQ2)

::: fill-box
**Research Question 2:** Do average standardized Medicare payment gaps differ
significantly across provider specialties in Massachusetts, and which specific
specialties drive the largest gaps?

**Why one-way ANOVA?** The outcome variable `log(payment_gap)` is continuous and
the grouping variable `Rndrng_Prvdr_Type` is categorical with multiple levels. ANOVA
tests whether the group means differ simultaneously rather than through a series of
pairwise comparisons, which would inflate the Type I error rate. The analysis is
restricted to the top 10 specialties by provider count to ensure adequate group sizes
for assumption checking and post-hoc inference.

**Hypotheses:**

$H_0$: The mean log payment gap is equal across all ten specialty groups.

$H_1$: At least one specialty group has a different mean log payment gap.
:::

```{r p2-setup}

top10_specialties <- ma_providers %>%
  count(Rndrng_Prvdr_Type, sort = TRUE) %>%
  head(10) %>%
  pull(Rndrng_Prvdr_Type)

ma_top10 <- ma_providers %>%
  filter(Rndrng_Prvdr_Type %in% top10_specialties) %>%
  mutate(Rndrng_Prvdr_Type = factor(Rndrng_Prvdr_Type))
```

------------------------------------------------------------------------

## Step 1: Assumption checks

::: why-box
**Why check assumptions before running ANOVA?**

Standard one-way ANOVA assumes (1) independence of observations, (2) normality of
residuals within each group, and (3) homogeneity of variance across groups. With
group sizes ranging from 1,033 to 5,496, the Central Limit Theorem ensures the
normality assumption is met asymptotically. The equal variance assumption is the
critical one to verify, as the boxplot already suggests substantially different
within-group spreads across specialties.
:::

```{r p2-levene}
cat("── Levene's Test for Homogeneity of Variance ──\n")
leveneTest(log(payment_gap) ~ Rndrng_Prvdr_Type, data = ma_top10)
```

::: pattern-box
**Interpretation:** Levene's test returns $F(9, 21{,}743) = 110.08$,
$p < 2.2 \times 10^{-16}$. The equal variance assumption is decisively violated.
Standard one-way ANOVA is therefore not appropriate; Welch's ANOVA is used instead,
as it does not assume homogeneity of variance across groups.
:::

------------------------------------------------------------------------

## Step 2: Normality check

::: why-box
**Why check normality?**

One-way ANOVA assumes that residuals within each group are approximately normally
distributed. With group sizes ranging from 1,033 to 5,496, the Central Limit Theorem
guarantees that the sampling distribution of group means is approximately normal
regardless of the underlying distribution, making the F-test asymptotically valid.
Shapiro-Wilk is nonetheless run on a random sample of residuals per group for
completeness. At these sample sizes, the test has sufficient power to detect trivially
small deviations from normality, so a significant result does not necessarily indicate
a practically meaningful violation.
:::

```{r p2-normality}
cat("── Shapiro-Wilk normality test per specialty group (sample of 500) ──\n")
ma_top10 %>%
  mutate(log_gap = log(payment_gap)) %>%
  group_by(Rndrng_Prvdr_Type) %>%
  summarise(
    n        = n(),
    sw_stat  = shapiro.test(sample(log_gap, min(500, n())))$statistic,
    sw_p     = shapiro.test(sample(log_gap, min(500, n())))$p.value,
    .groups  = "drop"
  ) %>%
  mutate(across(c(sw_stat, sw_p), ~round(., 4))) %>%
  arrange(sw_p) %>%
  print()
```

------------------------------------------------------------------------

## Step 3: Welch's ANOVA

::: why-box
**Why Welch's ANOVA instead of standard ANOVA?**

Standard one-way ANOVA pools the within-group variance into a single estimate,
which is unreliable when group variances differ substantially. Welch's ANOVA
estimates each group's variance separately, producing an F-statistic that is robust
to heteroscedasticity. The tradeoff is a reduction in degrees of freedom via the
Welch-Satterthwaite approximation, which makes the test slightly more conservative.
:::

```{r p2-welch-anova}
cat("── Welch's One-Way ANOVA ──\n")
welch_result <- oneway.test(
  log(payment_gap) ~ Rndrng_Prvdr_Type,
  data      = ma_top10,
  var.equal = FALSE
)
print(welch_result)

cat("\n── Classic ANOVA Table (for eta-squared and display) ──\n")
anova_classic <- aov(log(payment_gap) ~ Rndrng_Prvdr_Type, data = ma_top10)
summary(anova_classic)
```

------------------------------------------------------------------------

## Step 3: Effect size

::: why-box
**Why report effect size?**

A statistically significant F-test only indicates that at least one group mean
differs — it says nothing about the magnitude of the difference. Eta-squared
($\eta^2$) quantifies the proportion of total variance in log payment gap explained
by specialty group membership, providing a scale-independent measure of practical
significance. By convention, $\eta^2 = 0.01$ is small, $0.06$ is medium, and
$0.14$ is large.
:::

```{r p2-effect-size}
cat("── Effect size (eta-squared) ──\n")
anova_classic <- aov(log(payment_gap) ~ Rndrng_Prvdr_Type, data = ma_top10)
eta_sq <- sum(summary(anova_classic)[[1]]["Sum Sq"][1, ]) /
          sum(summary(anova_classic)[[1]]["Sum Sq"])
cat("Eta-squared:", round(eta_sq, 4), "\n")
```

------------------------------------------------------------------------

## Step 4: Post-hoc pairwise comparisons

::: why-box
**Why Games-Howell instead of Tukey HSD?**

Tukey's HSD assumes equal group variances, which Levene's test has rejected.
Games-Howell is the appropriate post-hoc procedure when variances are unequal: it
uses group-specific variance estimates and applies the Welch-Satterthwaite
approximation to degrees of freedom, controlling the familywise error rate without
assuming homoscedasticity.
:::

```{r p2-games-howell}
cat("── Games-Howell Post-Hoc Pairwise Comparisons ──\n")
gh_result <- ma_top10 %>%
  mutate(log_gap = log(payment_gap)) %>%
  games_howell_test(log_gap ~ Rndrng_Prvdr_Type) %>%
  arrange(p.adj) %>%
  print(n = Inf)
```

```{r p2-games-howell-sig}
cat("── Significant pairwise differences (p.adj < 0.05) ──\n")
gh_result %>%
  filter(p.adj < 0.05) %>%
  arrange(p.adj) %>%
  print(n = Inf)

cat("\n── Non-significant pairwise differences (p.adj >= 0.05) ──\n")
gh_result %>%
  filter(p.adj >= 0.05) %>%
  print(n = Inf)
```

```{r p2-cld}

cat("── Compact Letter Display ──\n")
# Extract p-values into named vector for multcompLetters
gh_wide <- gh_result %>%
  mutate(pair = paste(group1, group2, sep = "-")) %>%
  select(pair, p.adj)

p_vec <- setNames(gh_wide$p.adj, gh_wide$pair)
cld <- multcompLetters(p_vec)
print(cld)

# Tidy version joined to group means
cld_df <- data.frame(
  Rndrng_Prvdr_Type = names(cld$Letters),
  cld_letter        = cld$Letters
)

ma_top10 %>%
  group_by(Rndrng_Prvdr_Type) %>%
  summarise(mean_log = round(mean(log(payment_gap)), 3), .groups = "drop") %>%
  left_join(cld_df, by = "Rndrng_Prvdr_Type") %>%
  arrange(desc(mean_log)) %>%
  print()
```

::: why-box
**Why complement Welch's ANOVA with Kruskal-Wallis and Dunn's test?**
Welch's ANOVA and Games--Howell post-hoc tests are robust to unequal
variances --- the primary assumption violation identified by Levene's
test --- but both remain parametric procedures that model the
relationship between group means. The payment gap distribution within
each specialty is heavily right-skewed and heavy-tailed, meaning that
specialty-level means summarise skewed distributions rather than
symmetric ones. While the Central Limit Theorem ensures the F-test is
asymptotically valid at group sizes of 1,033--5,496, the concern is
not inferential validity but representativeness: do the mean-based
comparisons reflect what is actually happening in the bulk of each
specialty's distribution?

Kruskal-Wallis with Dunn's post-hoc test addresses this by operating
on ranks rather than raw values. It asks whether the \textit{distributions}
of payment gaps differ across specialties rather than whether the means
differ, making no assumptions about the shape of those distributions.
Dunn's test then performs pairwise comparisons on the same ranks with
Bonferroni correction for multiple comparisons.

If the two approaches agree --- the same specialty pairs emerge as
significantly different under both parametric and non-parametric
testing --- that convergence substantially strengthens confidence in
the RQ2 findings. If they disagree, that flags pairs where the
mean-based result may be driven by distributional shape rather than
a genuine location difference.
:::

::: fill-box
**Kruskal-Wallis and Dunn's test: what they do**

The Kruskal-Wallis test is the non-parametric equivalent of one-way
ANOVA. All observations across all groups are ranked from smallest to
largest, and the test asks whether the average ranks differ
significantly across specialty groups. Under the null hypothesis that
all specialties have the same payment gap distribution, the average
ranks should be roughly equal. A significant result indicates that at
least one specialty's distribution is shifted relative to the others.

Dunn's test then performs all 45 pairwise comparisons using those same
ranks. Bonferroni correction is applied to control the familywise
error rate across 45 simultaneous tests --- the same multiple
comparison concern that motivated Games--Howell in the parametric
analysis. A pair flagged as significant under Dunn's test has a
rank-based distribution difference that is unlikely to be due to
chance after correction.

**What to look for:** the non-significant pairs from Games--Howell
--- Family Practice vs.\ Physical Therapist and Family Practice vs.\
Physician Assistant --- should remain non-significant under Dunn's
test if the parametric and non-parametric approaches are telling the
same story. Any new non-significant pairs under Dunn's test, or any
pairs that flip from non-significant to significant, would indicate
that the mean-based result was sensitive to distributional shape.
:::

```{r p2-kruskal-wallis}
cat("── Kruskal-Wallis Test ──\n")
kw_result <- kruskal.test(
  payment_gap ~ Rndrng_Prvdr_Type,
  data = ma_top10
)
print(kw_result)

cat("\n── Dunn's Post-Hoc Test (Bonferroni correction) ──\n")
dunn_result <- dunn.test(
  x     = ma_top10$payment_gap,
  g     = ma_top10$Rndrng_Prvdr_Type,
  method = "bonferroni",
  alph   = 0.05,
  kw     = FALSE,
  label  = TRUE,
  wrap   = FALSE
)
```

```{r p2-kruskal-comparison}
# Tidy Dunn results and compare to Games-Howell
dunn_df <- data.frame(
  comparison = dunn_result$comparisons,
  p_adj      = dunn_result$P.adjusted
) %>%
  mutate(
    sig_dunn = p_adj < 0.05
  ) %>%
  arrange(p_adj)

cat("── Non-significant pairs under Dunn's test ──\n")
dunn_df %>%
  filter(!sig_dunn) %>%
  print()

cat("\n── Significant pairs: Dunn's test ──\n")
cat("Significant pairs:", sum(dunn_df$sig_dunn), "of", nrow(dunn_df), "\n")

cat("\n── Convergence check: Welch ANOVA vs Kruskal-Wallis ──\n")
cat("Welch's ANOVA p-value:    < 2.2e-16\n")
cat("Kruskal-Wallis p-value:  ", format(kw_result$p.value, scientific = TRUE), "\n")
cat("\nIf both reject H0 and non-significant pairs agree,\n")
cat("parametric and non-parametric approaches converge.\n")
```

::: pattern-box
**What the convergence check reveals:**
The Kruskal-Wallis test decisively rejects the null hypothesis that
payment gap distributions are equal across all ten specialties
($\chi^2(9) = 5285$, $p < 2.2 \times 10^{-16}$), consistent with
Welch's ANOVA ($F(9, 6{,}181.8) = 929.99$, $p < 2.2 \times 10^{-16}$).
The omnibus results from both approaches are in complete agreement.

Dunn's post-hoc tests with Bonferroni correction find 42 of 45 pairs
significantly different, compared to 43 of 45 under Games--Howell.
The two approaches agree on the core middle-tier finding: Family
Practice vs.\ Physical Therapist remains non-significant under both
methods. Two additional pairs are non-significant under Dunn's test
but significant under Games--Howell: Physical Therapist vs.\
Physician Assistant, and Anesthesiology vs.\ Diagnostic Radiology.

The Physical Therapist vs.\ Physician Assistant pair is
unsurprising given their similar medians (\$64,477 vs.\ \$57,653);
the rank-based test simply applies a more conservative criterion
than the mean-based one. The Anesthesiology vs.\ Diagnostic Radiology
discrepancy is more substantive: Games--Howell detects a difference
in means (\$266k vs.\ \$359k) that Dunn's test does not confirm at
the distributional level, suggesting the mean difference between
these two high-gap specialties is partly driven by the extreme right
tail of Diagnostic Radiology rather than a systematic shift in the
bulk of the distribution.

Overall the two approaches converge on the same structural picture:
a clear high-gap tier (Diagnostic Radiology, Anesthesiology, Emergency
Medicine), a low-gap tier (Licensed Clinical Social Worker, Psychiatry),
and a middle cluster whose internal boundaries are sensitive to whether
mean-based or rank-based comparisons are used. The core specialty
ranking is robust; the precise boundaries within the middle tier are not.
:::
------------------------------------------------------------------------

## Step 5: Group means plot

```{r p2-means-plot, fig.cap="Mean log payment gap with 95% confidence intervals by provider specialty. Specialties are ordered by mean log payment gap. Error bars reflect within-group standard error."}
ma_top10 %>%
  group_by(Rndrng_Prvdr_Type) %>%
  summarise(
    mean_log = mean(log(payment_gap)),
    se_log   = sd(log(payment_gap)) / sqrt(n()),
    .groups  = "drop"
  ) %>%
  ggplot(aes(
    x    = reorder(Rndrng_Prvdr_Type, mean_log),
    y    = mean_log,
    ymin = mean_log - 1.96 * se_log,
    ymax = mean_log + 1.96 * se_log
  )) +
  geom_pointrange(colour = "#2166AC", size = 0.7) +
  coord_flip() +
  labs(
    title    = "Mean Log Payment Gap by Provider Specialty",
    subtitle = "Massachusetts Medicare Providers, 2023 — Top 10 specialties by count",
    x        = NULL,
    y        = "Mean log(Payment Gap $)"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13),
    plot.subtitle = element_text(size = 10)
  )

ggsave(
  here::here("docs", "figures", "specialty_means.png"),
  width = 7, height = 5, dpi = 300
)
```

```{r p2-significance}
gh_result %>%
  filter(p.adj >= 0.05) %>%
  nrow()

gh_result %>%
  filter(p.adj >= 0.05) %>%
  select(group1, group2, estimate, p.adj)
```

::: why-box
**Why visualize all 97 specialties rather than just the top 10?**
The formal ANOVA is restricted to the top 10 specialties by provider
count to ensure adequate group sizes for reliable inference. This is
a principled analytical choice, but it raises a legitimate question:
do the 87 excluded specialties exhibit more extreme payment gaps than
those analyzed? If so, the ANOVA findings characterize only a subset
of the specialty landscape and the true upper end of the gap
distribution is underrepresented.

This visualization addresses that question directly by computing the
median payment gap for all 97 Massachusetts provider specialties and
ranking them from highest to lowest. No formal inference is performed
on the excluded specialties --- their group sizes range from 1 to 564
and many are too small for reliable Welch's ANOVA inference. The goal
is descriptive: to place the ANOVA findings in the context of the full
specialty landscape and identify whether any excluded specialties
occupy dramatically different positions in the gap distribution.

Two plots are presented. The first uses a log scale to show all 97
specialties simultaneously, revealing the full shape of the
distribution including facility and equipment-based outliers. The
second zooms into the individual provider range (median gap below
\$1M) where the ANOVA specialties are visible alongside the excluded
individual-type providers.
:::

::: fill-box
**Reading the full-spectrum specialty plots:**

**Log scale plot:** each point is one specialty. The y-axis is
logarithmic, meaning equal vertical distances represent equal
multiplicative differences rather than equal dollar differences.
Red points are the 10 specialties included in the formal ANOVA;
grey points are the 87 excluded specialties. The dashed horizontal
line marks the Massachusetts provider median (\$89,292). Points
above this line have above-median gaps; points below have
below-median gaps. The smooth decay from left to right shows how
the gap distribution tapers across the full specialty landscape.

**Zoomed plot:** the same data filtered to specialties with median
gaps below \$1M, which removes the facility and equipment-based
outliers and makes the individual provider structure visible. Blue
labels identify excluded specialties that rank above the highest
ANOVA specialty (Diagnostic Radiology, \$358,812). Any grey points
sitting above the leftmost red point (Diagnostic Radiology) are
specialties whose median gaps exceed the top ANOVA specialty despite
being excluded from formal testing due to low provider counts.

**What to look for:** how many excluded specialties rank above
Diagnostic Radiology, what types of specialties they are, and
whether their presence changes the interpretation of the ANOVA
findings about which specialties drive the largest gaps.
:::

```{r p2-all-specialties-ranked, fig.cap="Median payment gap for all 97 Massachusetts provider specialties, ranked from highest to lowest. Red points indicate the top 10 specialties by provider count included in the formal ANOVA. Grey points show the 87 excluded specialties. Dashed line marks the Massachusetts provider median ($89,292)."}

library(ggrepel)

# Compute median gap and provider count for all 97 specialties
all_specialties <- ma_providers %>%
  group_by(Rndrng_Prvdr_Type) %>%
  summarise(
    n          = n(),
    median_gap = median(payment_gap),
    mean_gap   = mean(payment_gap),
    p25_gap    = quantile(payment_gap, 0.25),
    p75_gap    = quantile(payment_gap, 0.75),
    .groups    = "drop"
  ) %>%
  arrange(desc(median_gap)) %>%
  mutate(
    rank        = row_number(),
    n_ranks     = n(),
    in_top10    = Rndrng_Prvdr_Type %in% top10_specialties,
    is_facility = rank <= 5 & !in_top10,
    is_bottom_interesting = Rndrng_Prvdr_Type %in% c(
      "Mass Immunizer Roster Biller",
      "Centralized Flu",
      "Registered Dietitian or Nutrition Professional",
      "Certified Nurse Midwife"
    )
  )

# Print bottom 10 for inspection
cat("── Bottom 10 specialties by median payment gap (all 97) ──\n")
all_specialties %>%
  arrange(median_gap) %>%
  select(rank, Rndrng_Prvdr_Type, n, median_gap) %>%
  head(10) %>%
  mutate(median_gap = scales::dollar(round(median_gap, 0))) %>%
  print()

# ── Plot 1: Log scale — full picture ─────────────────────────────────────────
p_full <- ggplot(all_specialties,
       aes(x = rank, y = median_gap, colour = in_top10)) +
  geom_point(
    aes(size = in_top10),
    alpha = 0.8
  ) +
  geom_hline(
    yintercept = median(ma_providers$payment_gap),
    linetype   = "dashed",
    colour     = "grey40",
    linewidth  = 0.6
  ) +
  # Top 10 ANOVA specialties — red labels
  geom_text_repel(
    data = all_specialties %>% filter(in_top10),
    aes(label = Rndrng_Prvdr_Type),
    size          = 2.5,
    colour        = "#C8102E",
    max.overlaps  = 20,
    box.padding   = 0.4,
    segment.color = "grey60"
  ) +
  # Top 5 facility outliers — blue labels
  geom_text_repel(
    data = all_specialties %>% filter(is_facility),
    aes(label = Rndrng_Prvdr_Type),
    size          = 2.5,
    colour        = "#2166AC",
    max.overlaps  = 20,
    box.padding   = 0.4,
    segment.color = "grey60"
  ) +
  # Bottom 10 — dark green labels
  geom_text_repel(
    data = all_specialties %>% filter(is_bottom_interesting),
    aes(label = paste0(Rndrng_Prvdr_Type, "\n(n=", n, ")")),
    size          = 2.3,
    colour        = "#1B7837",
    max.overlaps  = 20,
    box.padding   = 0.4,
    segment.color = "grey60",
    nudge_x       = -5,
    nudge_y       = -0.5
  ) +
  scale_colour_manual(
    values = c("TRUE" = "#C8102E", "FALSE" = "grey60"),
    labels = c("TRUE" = "Top 10 by count (ANOVA)", "FALSE" = "Excluded specialties"),
    name   = NULL
  ) +
  scale_size_manual(
    values = c("TRUE" = 2.5, "FALSE" = 1.2),
    guide  = "none"
  ) +
  scale_y_log10(labels = scales::dollar) +
  labs(
    title    = "Medicare Payment Gap Across All 97 Massachusetts Provider Specialties",
    subtitle = "Log scale — dashed line = Massachusetts provider median ($89,292)",
    x        = "Specialty rank (1 = highest median gap)",
    y        = "Median payment gap (log scale, $)"
  ) +
  theme_minimal() +
  theme(
    plot.title      = element_text(size = 13),
    plot.subtitle   = element_text(size = 10, colour = "grey40"),
    legend.position = "bottom"
  )

# ── Plot 2: Zoomed — individual providers only (median gap < $1M) ─────────────
p_zoom <- all_specialties %>%
  filter(median_gap < 1000000) %>%
  ggplot(aes(x = rank, y = median_gap, colour = in_top10)) +
  geom_point(
    aes(size = in_top10),
    alpha = 0.8
  ) +
  geom_hline(
    yintercept = median(ma_providers$payment_gap),
    linetype   = "dashed",
    colour     = "grey40",
    linewidth  = 0.6
  ) +
  # Top 10 ANOVA specialties — red labels
  geom_text_repel(
    data = all_specialties %>%
      filter(median_gap < 1000000, in_top10),
    aes(label = Rndrng_Prvdr_Type),
    size          = 2.5,
    colour        = "#C8102E",
    max.overlaps  = 20,
    box.padding   = 0.4,
    segment.color = "grey60"
  ) +
  # Excluded specialties above Diagnostic Radiology — blue labels
  geom_text_repel(
    data = all_specialties %>%
      filter(median_gap < 1000000, !in_top10, rank <= 15),
    aes(label = Rndrng_Prvdr_Type),
    size          = 2.5,
    colour        = "#2166AC",
    max.overlaps  = 20,
    box.padding   = 0.4,
    segment.color = "grey60"
  ) +
  # Bottom 10 — dark green labels
  geom_text_repel(
    data = all_specialties %>%
      filter(median_gap < 1000000, is_bottom_interesting),
    aes(label = paste0(Rndrng_Prvdr_Type)),
    size          = 2.3,
    colour        = "#1B7837",
    max.overlaps  = 20,
    box.padding   = 0.4,
    segment.color = "grey60",
    nudge_x       = -3,
    nudge_y       = 120000
  ) +
  scale_colour_manual(
    values = c("TRUE" = "#C8102E", "FALSE" = "grey60"),
    labels = c("TRUE" = "Top 10 by count (ANOVA)", "FALSE" = "Excluded specialties"),
    name   = NULL
  ) +
  scale_size_manual(
    values = c("TRUE" = 2.5, "FALSE" = 1.2),
    guide  = "none"
  ) +
  scale_y_continuous(labels = scales::dollar) +
  labs(
    title    = "Medicare Payment Gap — Individual Provider Specialties (Median Gap < $1M)",
    subtitle = "Zoomed view excluding facility/equipment-based outliers — dashed line = provider median ($89,292)",
    x        = "Specialty rank (1 = highest median gap)",
    y        = "Median payment gap ($)"
  ) +
  theme_minimal() +
  theme(
    plot.title      = element_text(size = 13),
    plot.subtitle   = element_text(size = 10, colour = "grey40"),
    legend.position = "bottom"
  )

print(p_full)
print(p_zoom)

ggsave(
  here::here("docs", "figures", "all_specialties_ranked_log.png"),
  plot  = p_full,
  width = 10, height = 6, dpi = 300
)

ggsave(
  here::here("docs", "figures", "all_specialties_ranked_zoom.png"),
  plot  = p_zoom,
  width = 10, height = 6, dpi = 300
)
```

::: pattern-box
**What the full-spectrum visualization reveals:**

The log scale plot confirms that the gap distribution follows a
smooth power-law decay across all 97 specialties. The top five
positions are occupied entirely by facility and equipment-based
entities --- Radiation Therapy Center (\$17.8M), Ambulatory Surgical
Center (\$3.0M), Mammography Center (\$2.0M), Portable X-Ray
Supplier (\$1.4M), and Independent Diagnostic Testing Facility
(\$1.3M) --- whose gaps are one to two orders of magnitude larger
than any individual physician specialty. These are not individual
practitioners but institutional billers operating under a different
billing logic, and their exclusion from the ANOVA is appropriate.

The zoomed plot reveals a more substantive finding: at least nine
excluded individual-type specialties have median gaps exceeding
Diagnostic Radiology (\$358,812), the highest-gap ANOVA specialty.
These include Ambulance Service Provider (\$867k), Hand Surgery
(\$622k), Interventional Pain Management (\$606k), Ophthalmology
(\$579k), Clinical Cardiac Electrophysiology (\$547k), Urology
(\$525k), Orthopedic Surgery (\$496k), and Interventional Cardiology
(\$496k). All are procedural or equipment-intensive specialties,
consistent with the ANOVA finding that procedure-intensive practice
patterns drive the largest gaps.

Two implications follow. First, the ANOVA limitation stated in
Section~5.2 --- that low-volume subspecialties may exhibit more
extreme gaps than those analyzed --- is confirmed. Diagnostic
Radiology anchors the top of the gap distribution among high-count
specialties, but in the full specialty landscape it ranks
approximately 20th. Second, the substantive conclusion of RQ2
is strengthened rather than contradicted: the specialties with the
largest gaps are uniformly procedural and equipment-intensive,
regardless of whether they were formally tested. The ANOVA findings
generalise directionally to the broader specialty landscape even
though they cannot be extended statistically beyond the top 10.
:::
------------------------------------------------------------------------

# Part 3: Logistic Regression (RQ3)

::: fill-box
**Research Question 3:** What provider and patient characteristics predict the
log-odds of a Massachusetts provider having an above-median standardized Medicare
payment gap?

**Why logistic regression?** The outcome variable \texttt{high\_gap} is binary (1 =
above-median payment gap, 0 = at or below median). Logistic regression models the
log-odds of the outcome as a linear combination of predictors, allowing each
coefficient to be interpreted as the change in log-odds for a one-unit increase in
that predictor, holding all others constant. Exponentiating coefficients yields odds
ratios (OR), which express the multiplicative change in the odds of being a high-gap
provider.

**Hypotheses:**

$H_0$: None of the six predictors are associated with the log-odds of being a
high-gap provider.

$H_1$: At least one predictor is significantly associated with the log-odds of
being a high-gap provider.
:::

```{r p3-logistic-fit}
model_logistic <- glm(
  high_gap ~ log(Tot_Benes) + log(srvcs_per_bene) + Bene_Avg_Risk_Scre +
             pct_dual + drug_provider + log(Tot_HCPCS_Cds),
  data   = ma_providers_rq3,
  family = binomial(link = "logit")
)

cat("── Logistic Regression Summary ──\n")
summary(model_logistic)
```

```{r p3-logistic-standardized}
model_parameters(model_logistic, exponentiate = TRUE, standardize = "refit") |>
  print()
```

------------------------------------------------------------------------

## Odds Ratios and Confidence Intervals

::: why-box
**Why report odds ratios?**

Logistic regression coefficients are on the log-odds scale, which is not directly
interpretable. Exponentiating each coefficient yields an odds ratio --- the
multiplicative factor by which the odds of being a high-gap provider change for a
one-unit increase in the predictor, holding all other variables constant. An OR
above 1 indicates higher odds of high-gap classification; an OR below 1 indicates
lower odds.
:::

```{r p3-odds-ratios}
cat("── Odds Ratios with 95% Confidence Intervals ──\n")
tidy(model_logistic, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  print()
```

------------------------------------------------------------------------

## Model Fit

```{r p3-model-fit}
cat("── Null and Residual Deviance ──\n")
cat("Null deviance:     ", round(model_logistic$null.deviance, 2),
    "on", model_logistic$df.null, "df\n")
cat("Residual deviance: ", round(model_logistic$deviance, 2),
    "on", model_logistic$df.residual, "df\n")

cat("\n── McFadden Pseudo-R² ──\n")
mcfadden <- 1 - (model_logistic$deviance / model_logistic$null.deviance)
cat("McFadden R²:", round(mcfadden, 4), "\n")

cat("\n── AIC ──\n")
cat("AIC:", round(AIC(model_logistic), 2), "\n")
```

------------------------------------------------------------------------

## Threshold Selection and Confusion Matrix

::: why-box
**Why not use 0.5 as the classification threshold?**

The default threshold of 0.5 assumes equal costs for false positives and false
negatives, and implicitly assumes balanced classes. In this dataset, 62\% of
providers are high-gap classifiers, so a naive model predicting high-gap for
everyone would already achieve 62\% accuracy. More importantly, the research
context favors sensitivity over specificity: correctly identifying high-gap
providers is more valuable than avoiding false alarms, as the goal is to flag
providers warranting further review. The threshold is selected by examining the
ROC curve and choosing the value that maximises the sum of sensitivity and
specificity (Youden's J statistic).
:::

```{r p3-threshold}

pred_probs <- predict(model_logistic, type = "response")

roc_obj <- roc(ma_providers_rq3$high_gap, pred_probs, quiet = TRUE)

cat("── ROC AUC ──\n")
cat("AUC:", round(auc(roc_obj), 4), "\n")

# Youden's J: maximise sensitivity + specificity - 1
youden_idx <- which.max(roc_obj$sensitivities + roc_obj$specificities - 1)
optimal_threshold <- roc_obj$thresholds[youden_idx]
cat("\n── Optimal threshold (Youden's J) ──\n")
cat("Threshold:", round(optimal_threshold, 4), "\n")
```

```{r p3-confusion-matrix}
pred_class <- ifelse(pred_probs >= optimal_threshold, 1, 0)

cat("── Confusion Matrix ──\n")
cm <- table(Predicted = pred_class, Actual = ma_providers_rq3$high_gap)
print(cm)

TP <- cm[2, 2]; TN <- cm[1, 1]
FP <- cm[2, 1]; FN <- cm[1, 2]

accuracy    <- round((TP + TN) / sum(cm), 4)
sensitivity <- round(TP / (TP + FN), 4)
specificity <- round(TN / (TN + FP), 4)
precision   <- round(TP / (TP + FP), 4)

cat("\n── Classification Metrics ──\n")
cat("Accuracy:   ", accuracy,    "\n")
cat("Sensitivity:", sensitivity, "\n")
cat("Specificity:", specificity, "\n")
cat("Precision:  ", precision,   "\n")
```

------------------------------------------------------------------------

## ROC Curve

```{r p3-roc-plot, fig.cap="ROC curve for the logistic regression model predicting high-gap provider classification. The dashed diagonal represents a random classifier (AUC = 0.5). The filled point marks the optimal threshold selected by Youden's J statistic."}
roc_df <- data.frame(
  fpr  = 1 - roc_obj$specificities,
  tpr  = roc_obj$sensitivities
)

optimal_point <- data.frame(
  fpr = 1 - roc_obj$specificities[youden_idx],
  tpr = roc_obj$sensitivities[youden_idx]
)

ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(colour = "#2166AC", linewidth = 0.9) +
  geom_abline(linetype = "dashed", colour = "grey50") +
  geom_point(
    data   = optimal_point,
    aes(x = fpr, y = tpr),
    colour = "#C8102E", size = 3
  ) +
  annotate(
    "text",
    x     = optimal_point$fpr + 0.05,
    y     = optimal_point$tpr - 0.03,
    label = paste0("Threshold = ", round(optimal_threshold, 3)),
    size  = 3.5, colour = "#C8102E"
  ) +
  labs(
    title    = "ROC Curve — High-Gap Provider Classification",
    subtitle = paste0("AUC = ", round(auc(roc_obj), 4),
                      "  |  n = ", scales::comma(nrow(ma_providers_rq3))),
    x        = "False Positive Rate (1 - Specificity)",
    y        = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13),
    plot.subtitle = element_text(size = 10, colour = "grey40")
  )

ggsave(
  here::here("docs", "figures", "roc_curve.png"),
  width = 6, height = 5, dpi = 300
)
```

::: why-box
**Why run a sensitivity analysis on the classification threshold?**
The \texttt{high\_gap} outcome is defined at the Massachusetts
provider median (\$89,292) --- a convenient and interpretable cutoff,
but an arbitrary one. There is no clinical or policy reason why the
50th percentile specifically defines "high" billing behavior. A
provider at the 51st percentile and one at the 95th percentile are
treated identically as high-gap, while a provider at the 49th
percentile is treated as low-gap despite being nearly identical to
the 51st.

If the logistic regression findings are genuinely capturing
structural differences in billing behavior, the odds ratios should
remain stable when the outcome definition shifts. If a finding
disappears or reverses at a different threshold --- for example, if
the dual-eligibility association vanishes when only the top quarter
of providers are classified as high-gap --- that would suggest the
result is sensitive to the specific cutoff rather than reflecting a
robust underlying pattern.

The model is therefore refit at four thresholds: the 40th, 50th,
60th, and 75th percentiles of the payment gap distribution in the
RQ3 analysis dataset. All six predictors and the full model
specification are held constant; only the definition of the outcome
variable changes. This is not a different analysis --- it is the
same analysis asking whether the same predictors remain reliable
signals of high-gap classification when "high-gap" is defined more
or less strictly.
:::

```{r p3-threshold-sensitivity}
cat("── Sensitivity Analysis: Odds Ratios Across Classification Thresholds ──\n")

# Define thresholds at 40th, 50th, 60th, and 75th percentiles
thresholds <- c(0.40, 0.50, 0.60, 0.75)

sensitivity_results <- bind_rows(
  lapply(thresholds, function(q) {
    
    # Redefine high_gap at each percentile threshold
    cutoff <- quantile(ma_providers_rq3$payment_gap, q)
    
    df <- ma_providers_rq3 %>%
      mutate(high_gap_q = as.integer(payment_gap > cutoff))
    
    # Refit logistic regression with new outcome
    fit <- glm(
      high_gap_q ~ log(Tot_Benes) + log(srvcs_per_bene) + Bene_Avg_Risk_Scre +
                   pct_dual + drug_provider + log(Tot_HCPCS_Cds),
      data   = df,
      family = binomial(link = "logit")
    )
    
    # Extract odds ratios
    tidy(fit, conf.int = TRUE, exponentiate = TRUE) %>%
      filter(term != "(Intercept)") %>%
      mutate(
        threshold     = q,
        cutoff_dollar = scales::dollar(round(cutoff, 0))
      )
  })
)

# Print clean summary table
sensitivity_results %>%
  select(threshold, cutoff_dollar, term, estimate, conf.low, conf.high, p.value) %>%
  mutate(across(c(estimate, conf.low, conf.high), ~round(., 3))) %>%
  mutate(p.value = round(p.value, 4)) %>%
  arrange(term, threshold) %>%
  print(n = Inf)
```

::: fill-box
**Reading the threshold sensitivity plot:**
Each panel shows one predictor. The x-axis shows the four
classification thresholds: P40 (40th percentile, \$95,769), P50
(50th percentile, \$132,121 --- the main analysis), P60 (60th
percentile, \$178,079), and P75 (75th percentile, \$292,776). The
y-axis shows the odds ratio at each threshold. The shaded band is
the 95\% confidence interval. The dashed horizontal line marks
OR = 1, which would indicate no association.

**What to look for:**

- **Flat lines staying well away from OR = 1** --- the predictor is
  robust. Its association with high-gap classification does not
  depend on how strictly high-gap is defined.
- **Lines drifting toward OR = 1** --- the predictor attenuates at
  stricter thresholds. It distinguishes above-median providers from
  below-median ones but loses power when only the most extreme
  billers are in the high-gap group.
- **Lines crossing OR = 1 or changing direction** --- the finding
  is threshold-sensitive and should be interpreted cautiously.
- **For predictors with OR below 1** (pct\_dual, drug\_provider):
  moving further below 1 means the negative association is becoming
  more pronounced; moving toward 1 means it is attenuating.

Note that the y-axis is free-scaled across panels --- do not compare
the absolute vertical range across panels, only the shape of each
line relative to its own OR = 1 reference.
:::

```{r p3-threshold-sensitivity-plot}

sensitivity_results <- sensitivity_results %>%
  mutate(
    term_label = case_when(
      term == "log(Tot_Benes)"      ~ "log(Tot_Benes)",
      term == "log(srvcs_per_bene)" ~ "log(srvcs_per_bene)",
      term == "Bene_Avg_Risk_Scre"  ~ "Bene_Avg_Risk_Scre",
      term == "pct_dual"            ~ "pct_dual",
      term == "drug_providerTRUE"   ~ "drug_provider",
      term == "drug_provider"       ~ "drug_provider",
      term == "log(Tot_HCPCS_Cds)"  ~ "log(Tot_HCPCS_Cds)",
      TRUE ~ term
    ),
    threshold_label = paste0("P", threshold * 100)
  )

ggplot(sensitivity_results,
       aes(x = threshold_label, y = estimate, group = term_label)) +
  # CI band
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    fill  = "#2166AC",
    alpha = 0.15
  ) +
  # Line connecting estimates
  geom_line(
    colour    = "#2166AC",
    linewidth = 0.8
  ) +
  # Points
  geom_point(
    colour = "#2166AC",
    size   = 3
  ) +
  # Reference line at OR = 1
  geom_hline(
    yintercept = 1,
    linetype   = "dashed",
    colour     = "grey40",
    linewidth  = 0.6
  ) +
  facet_wrap(~ term_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "Logistic Regression Odds Ratios Across Classification Thresholds",
    subtitle = "Shaded band = 95% CI. P50 = median threshold (main analysis).\nDashed line = OR of 1 (no effect).",
    x        = "Classification threshold",
    y        = "Odds ratio"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13),
    plot.subtitle = element_text(size = 9, colour = "grey40"),
    strip.text    = element_text(size = 10, face = "bold")
  )

ggsave(
  here::here("docs", "figures", "threshold_sensitivity.png"),
  width = 9, height = 6, dpi = 300
)
```

::: pattern-box
**What the sensitivity analysis reveals:**
All six predictors maintain the same sign and remain statistically
significant at every threshold. No finding reverses or approaches
the null. The logistic regression results are not an artifact of
the median split.

**\texttt{log(Tot\_Benes)}** is the most stable predictor: 6.89,
6.95, 6.99 at P40 through P60, attenuating modestly to 6.04 at P75.
Provider volume dominates high-gap classification at every threshold.

**\texttt{pct\_dual}** shows the most substantively interesting
pattern. The negative association becomes more pronounced as the
threshold rises --- OR = 0.202 at P40, dropping to 0.090 at P60,
then recovering slightly to 0.121 at P75. Providers serving panels
with higher proportions of dual-eligible patients are increasingly
unlikely to be classified as high-gap as the definition of high-gap
tightens. The OR remains far below 1 at every threshold and the
confidence bands never approach the null.

**\texttt{log(srvcs\_per\_bene)}** shows a mild upward trend from
2.66 at P40 to 2.83 at P75. Billing intensity is slightly more
discriminating at stricter thresholds, consistent with the Model 1
quantile regression finding that billing intensity matters more for
providers at the upper tail of the gap distribution.

**\texttt{Bene\_Avg\_Risk\_Scre}**, **\texttt{log(Tot\_HCPCS\_Cds)}**,
and **\texttt{drug\_provider}** all attenuate at P75 but remain
strongly significant. The consistent attenuation across these three
predictors at the strictest threshold suggests that the very top
quarter of the gap distribution is somewhat less differentiated by
patient complexity, practice breadth, and drug billing status ---
volume and billing intensity become the relatively more dominant
signals at the extreme top.
:::


::: why-box
**Why add specialty as a covariate to the logistic regression?**
The base logistic regression includes six predictors spanning
billing behavior, patient complexity, practice breadth, and
dual-eligibility burden, but not specialty. This is a
deliberate choice: specialty is not one of the six readily
observable characteristics the model is designed to use as a
screening tool, and adding specialty dummies would change the
question from "what provider characteristics predict high-gap
classification" to "what characteristics predict high-gap
classification within a given specialty."

However, omitting specialty raises a legitimate concern. The
observed associations, particularly for \texttt{pct\_dual},
could reflect specialty composition rather than independent
billing behavior. Providers serving high dual-eligibility panels
may simply be concentrated in inherently low-gap specialties
such as Psychiatry or Licensed Clinical Social Work, in which
case the dual-eligibility finding would be a specialty effect
in disguise rather than a genuine association with billing
behavior.

The specialty-controlled refit addresses this directly by adding
specialty group membership as a covariate (top 10 specialties
by provider count plus an \textit{Other} category) and comparing
the core odds ratios to those from the base model. If the
findings are robust, the odds ratios should be stable or
strengthen after specialty adjustment. If they attenuate toward
1, the base model findings were partially explained by specialty
composition.
:::

```{r p3-logistic-specialty}
# Add specialty as a covariate to the logistic regression
# Use top 10 specialties as explicit dummies, collapse rest to "Other"
ma_providers_rq3_specialty <- ma_providers_rq3 %>%
  mutate(
    specialty_group = ifelse(
      Rndrng_Prvdr_Type %in% top10_specialties,
      Rndrng_Prvdr_Type,
      "Other"
    ),
    specialty_group = factor(specialty_group)
  )

model_logistic_specialty <- glm(
  high_gap ~ log(Tot_Benes) + log(srvcs_per_bene) + Bene_Avg_Risk_Scre +
             pct_dual + drug_provider + log(Tot_HCPCS_Cds) +
             specialty_group,
  data   = ma_providers_rq3_specialty,
  family = binomial(link = "logit")
)

cat("── Logistic Regression with Specialty Controls ──\n")
summary(model_logistic_specialty)

cat("\n── McFadden Pseudo-R² with specialty ──\n")
mcfadden_sp <- 1 - (model_logistic_specialty$deviance /
                    model_logistic_specialty$null.deviance)
cat("McFadden R²:", round(mcfadden_sp, 4), "\n")

cat("\n── AUC with specialty ──\n")
pred_probs_sp <- predict(model_logistic_specialty, type = "response")
roc_sp <- roc(ma_providers_rq3_specialty$high_gap,
              pred_probs_sp, quiet = TRUE)
cat("AUC:", round(auc(roc_sp), 4), "\n")
```

::: fill-box
**Reading the specialty robustness comparison:**
The comparison plot shows odds ratios for the six core predictors
side by side: the base model (blue) and the specialty-controlled
model (red). The dashed horizontal line marks OR = 1 (no
effect). Each panel shows one predictor on a free y-axis scale.

**What to look for:**

- **Points close together** --- the predictor is robust to
  specialty adjustment. Its association with high-gap
  classification is not explained by specialty composition.
- **Red point closer to 1 than blue** --- the predictor
  attenuates after specialty control. Some of its apparent
  effect was a specialty proxy.
- **Red point further from 1 than blue** --- the predictor
  strengthens after specialty control. Specialty was acting
  as a suppressor, and the true within-specialty association
  is stronger than the base model suggested.

For predictors with OR below 1 (\texttt{pct\_dual},
\texttt{drug\_provider}): moving further below 1 means
the negative association strengthens; moving toward 1
means it attenuates.

The model fit comparison (McFadden $R^2$ and AUC) indicates
how much additional explanatory power specialty adds. A large
improvement would suggest the base model was substantially
misspecified; a modest improvement would confirm specialty
adds some but not decisive explanatory power.
:::

```{r p3-logistic-specialty-comparison}
# Compare key ORs between base model and specialty-controlled model
cat("── Key OR comparison: base vs specialty-controlled model ──\n")

# Core predictors to compare — exclude specialty dummies
core_terms <- c(
  "log(Tot_Benes)",
  "log(srvcs_per_bene)",
  "Bene_Avg_Risk_Scre",
  "pct_dual",
  "drug_provider",
  "log(Tot_HCPCS_Cds)"
)

bind_rows(
  tidy(model_logistic, conf.int = TRUE, exponentiate = TRUE) %>%
    mutate(model = "Base model"),
  tidy(model_logistic_specialty, conf.int = TRUE, exponentiate = TRUE) %>%
    mutate(model = "Specialty-controlled")
) %>%
  filter(term %in% core_terms) %>%
  select(model, term, estimate, conf.low, conf.high, p.value) %>%
  mutate(
    across(c(estimate, conf.low, conf.high), ~round(., 3)),
    p.value = round(p.value, 4)
  ) %>%
  arrange(term, model) %>%
  print(n = Inf)

cat("\n── Model fit comparison ──\n")
cat("Base model      — McFadden R²:", round(mcfadden, 4),
    "| AUC:", round(auc(roc_obj), 4), "\n")
cat("Specialty model — McFadden R²:", round(mcfadden_sp, 4),
    "| AUC:", round(auc(roc_sp), 4), "\n")
```


```{r p3-logistic-specialty-plot}

or_comparison <- bind_rows(
  tidy(model_logistic, conf.int = TRUE, exponentiate = TRUE) %>%
    mutate(model = "Base model"),
  tidy(model_logistic_specialty, conf.int = TRUE, exponentiate = TRUE) %>%
    mutate(model = "Specialty-controlled")
) %>%
  filter(term %in% core_terms) %>%
  mutate(
    term = case_when(
      term == "log(Tot_Benes)"      ~ "log(Tot_Benes)",
      term == "log(srvcs_per_bene)" ~ "log(srvcs_per_bene)",
      term == "Bene_Avg_Risk_Scre"  ~ "Bene_Avg_Risk_Scre",
      term == "pct_dual"            ~ "pct_dual",
      term == "drug_providerTRUE"   ~ "drug_provider",
      term == "drug_provider"       ~ "drug_provider",
      term == "log(Tot_HCPCS_Cds)"  ~ "log(Tot_HCPCS_Cds)",
      TRUE ~ term
    )
  )

ggplot(or_comparison,
       aes(x = model, y = estimate,
           ymin = conf.low, ymax = conf.high,
           colour = model)) +
  geom_pointrange(size = 0.6) +
  geom_hline(
    yintercept = 1,
    linetype   = "dashed",
    colour     = "grey40",
    linewidth  = 0.6
  ) +
  facet_wrap(~ term, scales = "free_y", ncol = 2) +
  scale_colour_manual(
    values = c(
      "Base model"           = "#2166AC",
      "Specialty-controlled" = "#C8102E"
    ),
    guide = "none"
  ) +
  labs(
    title    = "Logistic Regression OR Comparison: Base vs Specialty-Controlled",
    subtitle = "Blue = base model; Red = specialty dummies added. Dashed line = OR of 1.",
    x        = NULL,
    y        = "Odds ratio"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13),
    plot.subtitle = element_text(size = 10, colour = "grey40"),
    strip.text    = element_text(size = 10, face = "bold")
  )

ggsave(
  here::here("docs", "figures", "specialty_controlled_or.png"),
  width = 9, height = 6, dpi = 300
)
```

::: pattern-box
**What the specialty robustness check reveals:**
All six core findings are stable or strengthen after controlling
for specialty group membership. No predictor attenuates toward
the null.

**\texttt{pct\_dual}}** moves from OR = 0.212 to OR = 0.168
after specialty adjustment --- strengthening rather than
attenuating. This directly rules out the most obvious
alternative explanation for the dual-eligibility finding:
high dual-eligibility providers are not simply concentrated
in inherently low-gap specialties. The association between
dual-eligibility burden and lower gap odds persists and
intensifies within specialty.

**\texttt{log(Tot\_Benes)}}** strengthens from 6.72 to 7.73,
indicating that within a given specialty, volume is an even
stronger differentiator of billing behavior than it appears
across specialties. Specialty and volume are partially
correlated --- high-volume providers tend to cluster in
certain specialties --- and controlling for specialty isolates
the within-specialty volume effect.

**\texttt{log(Tot\_HCPCS\_Cds)}}** attenuates modestly from
3.77 to 3.62, the only predictor that moves toward the null.
This is expected: practice breadth and specialty are related,
since different specialties have characteristically different
procedure repertoires. The attenuation is modest, confirming
that breadth captures something beyond specialty membership.

**\texttt{log(srvcs\_per\_bene)}}**, **\texttt{Bene\_Avg\_Risk\_Scre}}**,
and **\texttt{drug\_provider}}** are all stable across
specifications.

McFadden $R^2$ improves from 0.439 to 0.462 and AUC from
0.905 to 0.913 after adding specialty --- a modest improvement
confirming that specialty adds real but not decisive
explanatory power. The base model was not substantially
misspecified; specialty is a useful additional signal but
not a confound of the core findings.
:::
------------------------------------------------------------------------

# Part 4: Bootstrap & Permutation Test (RQ4)

::: fill-box
**Research Question 4:** Does the median standardized Medicare payment gap differ
significantly between Massachusetts providers whose patient panels have low versus
high proportions of dual-eligible beneficiaries?
:::

::: why-box
**Why Bootstrap & Permutation Tests?**

The payment gap distribution is heavily right-skewed with extreme outliers, making
parametric confidence intervals for the median unreliable. The median is not amenable
to a simple formula-based standard error under these distributional conditions, and
even after log transformation the tails remain heavy. Bootstrap resampling constructs
a confidence interval for each group's median empirically, making no distributional
assumptions. The permutation test then assesses whether the observed difference in
median payment gaps between the two groups could plausibly arise by chance under the
null hypothesis of no difference, without assuming normality in either group.

The dual-eligibility comparison is motivated directly by RQ3, which identified
`pct_dual` as a strong negative predictor of high-gap classification (OR = 0.186).
RQ4 examines whether this association persists unconditionally and quantifies its
magnitude in raw dollar terms.
:::

```{r p4-setup}
set.seed(6105)

ma_providers_rq4 <- ma_providers_rq3 %>%
  mutate(
    dual_group = ifelse(
      pct_dual > median(pct_dual),
      "High dual-eligibility",
      "Low dual-eligibility"
    )
  )

cat("── Group sizes ──\n")
table(ma_providers_rq4$dual_group)

low_gap  <- ma_providers_rq4$payment_gap[
  ma_providers_rq4$dual_group == "Low dual-eligibility"
]
high_gap <- ma_providers_rq4$payment_gap[
  ma_providers_rq4$dual_group == "High dual-eligibility"
]

obs_diff <- median(low_gap) - median(high_gap)

cat("\n── Observed medians ──\n")
cat("Low dual-eligibility:  $", scales::comma(round(median(low_gap))), "\n")
cat("High dual-eligibility: $", scales::comma(round(median(high_gap))), "\n")
cat("Observed difference:   $", scales::comma(round(obs_diff)), "\n")
```

------------------------------------------------------------------------

## Step 1: Bootstrap Confidence Intervals for Each Group Median

::: why-box
**Why bootstrap the median separately for each group?**

The bootstrap constructs a confidence interval by repeatedly resampling each group
with replacement and computing the median on each resample. With 10,000 replicates,
the distribution of bootstrapped medians approximates the sampling distribution of
the observed median, from which percentile-based 95\% confidence intervals are
derived. No assumption is made about the shape of the payment gap distribution within
each group. Constructing separate CIs for each group's median makes the uncertainty
in each estimate visible before the formal hypothesis test.
:::

```{r p4-bootstrap}
B <- 10000

# Bootstrap CI for low dual-eligibility median
boot_low <- replicate(B, median(sample(low_gap, length(low_gap), replace = TRUE)))
ci_low   <- quantile(boot_low, c(0.025, 0.975))

# Bootstrap CI for high dual-eligibility median
boot_high <- replicate(B, median(sample(high_gap, length(high_gap), replace = TRUE)))
ci_high   <- quantile(boot_high, c(0.025, 0.975))

cat("── Bootstrap Results (B = 10,000) ──\n\n")

cat("Low dual-eligibility:\n")
cat("  Observed median: $", scales::comma(round(median(low_gap))), "\n")
cat("  95% CI:         [$", scales::comma(round(ci_low[1])),
    ", $", scales::comma(round(ci_low[2])), "]\n\n")

cat("High dual-eligibility:\n")
cat("  Observed median: $", scales::comma(round(median(high_gap))), "\n")
cat("  95% CI:         [$", scales::comma(round(ci_high[1])),
    ", $", scales::comma(round(ci_high[2])), "]\n")
```

```{r p4-bootstrap-plot, fig.cap="Bootstrap distributions of the median payment gap for low and high dual-eligibility providers (10,000 replicates each). Vertical solid lines mark the observed medians; dashed lines mark the 95% confidence interval bounds. Non-overlapping intervals indicate a statistically reliable difference."}
boot_df <- bind_rows(
  data.frame(group = "Low dual-eligibility",  median = boot_low),
  data.frame(group = "High dual-eligibility", median = boot_high)
)

ci_df <- data.frame(
  group  = c("Low dual-eligibility", "Low dual-eligibility",
             "High dual-eligibility", "High dual-eligibility"),
  x      = c(ci_low[1], ci_low[2], ci_high[1], ci_high[2]),
  ltype  = rep(c("lower", "upper"), 2)
)

obs_df <- data.frame(
  group = c("Low dual-eligibility", "High dual-eligibility"),
  x     = c(median(low_gap), median(high_gap))
)

ggplot(boot_df, aes(x = median, fill = group)) +
  geom_histogram(bins = 60, alpha = 0.8) +
  geom_vline(data = obs_df, aes(xintercept = x),
             colour = "#C8102E", linewidth = 1) +
  geom_vline(data = ci_df, aes(xintercept = x),
             colour = "#C8102E", linewidth = 0.7, linetype = "dashed") +
  scale_fill_manual(values = c(
    "Low dual-eligibility"  = "#C8102E",
    "High dual-eligibility" = "#2166AC"
  ), name = NULL) +
  scale_x_continuous(labels = scales::dollar) +
  facet_wrap(~group, scales = "free_x") +
  labs(
    title    = "Bootstrap Distribution of Median Payment Gap by Dual-Eligibility Group",
    subtitle = "B = 10,000 replicates | Solid line = observed median | Dashed = 95% CI",
    x        = "Bootstrapped Median ($)",
    y        = "Count"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title      = element_text(size = 13),
    plot.subtitle   = element_text(size = 10, colour = "grey40")
  )

ggsave(
  here::here("docs", "figures", "bootstrap_distribution.png"),
  width = 10, height = 5, dpi = 300
)
```

------------------------------------------------------------------------

## Step 2: Permutation Test

::: why-box
**Why a permutation test?**

The permutation test asks: if group membership were assigned randomly, how often
would we observe a difference in median payment gaps as large as the one actually
observed? Under the null hypothesis of no difference, the group labels are
exchangeable. By randomly permuting the labels 10,000 times and computing the
difference in medians each time, we build an empirical null distribution against
which the observed difference is compared. The p-value is the proportion of permuted
differences that equal or exceed the observed difference in absolute magnitude.

**Hypotheses:**

$H_0$: The median payment gap is equal for low and high dual-eligibility providers.

$H_1$: The median payment gap differs between low and high dual-eligibility providers.
:::

```{r p4-permutation}
N_perm <- 10000

perm_diffs <- replicate(N_perm, {
  shuffled <- sample(ma_providers_rq4$dual_group)
  median(ma_providers_rq4$payment_gap[shuffled == "Low dual-eligibility"]) -
  median(ma_providers_rq4$payment_gap[shuffled == "High dual-eligibility"])
})

perm_p <- mean(abs(perm_diffs) >= abs(obs_diff))

cat("── Permutation Test Results (N = 10,000) ──\n")
cat("Observed difference: $", scales::comma(round(obs_diff)), "\n")
cat("Permutation p-value:", ifelse(perm_p == 0, "< 0.0001", round(perm_p, 4)), "\n")
```

```{r p4-permutation-plot, fig.cap="Empirical null distribution of 10,000 permuted differences in median payment gap. The red vertical lines mark the observed difference and its negative mirror. The proportion of permuted differences falling outside these lines gives the two-sided permutation p-value."}
ggplot(data.frame(diff = perm_diffs), aes(x = diff)) +
  geom_histogram(bins = 60, fill = "#2166AC", alpha = 0.8) +
  geom_vline(xintercept =  obs_diff, colour = "#C8102E", linewidth = 1) +
  geom_vline(xintercept = -obs_diff, colour = "#C8102E", linewidth = 1) +
  scale_x_continuous(labels = scales::dollar) +
  labs(
    title    = "Permutation Null Distribution — Difference in Median Payment Gap",
    subtitle = paste0("N = 10,000 permutations | Observed difference = $",
                      scales::comma(round(obs_diff))),
    x        = "Permuted Difference in Medians ($)",
    y        = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(size = 13),
    plot.subtitle = element_text(size = 10, colour = "grey40")
  )

ggsave(
  here::here("docs", "figures", "permutation_distribution.png"),
  width = 7, height = 5, dpi = 300
)
```

------------------------------------------------------------------------

## Step 3: Parametric Comparison

::: why-box
**Why compare to a parametric alternative?**

The Wilcoxon rank-sum test is a nonparametric alternative to the two-sample t-test
that does not assume normality but relies on asymptotic approximations for large
samples. Comparing its conclusion to the permutation test result assesses whether
the two approaches agree. The Hodges-Lehmann estimate reported by the Wilcoxon test
with `conf.int = TRUE` provides a pseudomedian difference that can be compared
directly to the bootstrap CIs.
:::

```{r p4-wilcoxon}
cat("── Wilcoxon Rank-Sum Test ──\n")
wilcox.test(
  payment_gap ~ dual_group,
  data       = ma_providers_rq4,
  conf.int   = TRUE,
  conf.level = 0.95
)
```